---
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
urlcolor: black
linkcolor: black
bibliography: 99-refs.bibtex
fontsize: 12pt
geometry: margin = 1in
header-includes:
- \usepackage{placeins}
- \usepackage{fancyhdr}
- \usepackage{geometry}
- \usepackage{setspace}
- \usepackage{chngcntr}
- \usepackage{microtype}
- \counterwithin{figure}{section}
- \counterwithin{table}{section}
- \geometry{a4paper,top=3cm,bottom=3.5cm,left=2.5cm,right=2.5cm,bindingoffset=5mm}
- \onehalfspacing
---

---
nocite: | 
  @Example1999, @Example2000
...

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.path = 'figures/',
                      echo = FALSE, warning = FALSE, message = FALSE)
```

\pagenumbering{gobble}

```{r child = 'titlepage.Rmd'}
```

<!-- \newpage -->

<!-- ```{r child = 'declaration.Rmd'} -->
<!-- ``` -->

\newpage
\pagestyle{fancy}

\fancyhead[LE,RO]{}
\fancyhead[LO,RE]{}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0pt}

\pagenumbering{roman}

\fancyhead[CO,CE]{Abstract}
\section*{Abstract}
\addcontentsline{toc}{section}{Abstract}

```{r child = 'abstract.Rmd'}
```

\newpage
\fancyhead[CO,CE]{Acknowledgements}
\section*{Acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}

```{r child = 'acknowledgements.Rmd'}
```

\newpage
\fancyhead[CO,CE]{Table of Contents}
\setcounter{tocdepth}{2}
\tableofcontents

\newpage
\pagenumbering{arabic}

\newpage
\fancyhead[CO,CE]{Introduction}

```{r child = '1-introduction.Rmd'}
```

\FloatBarrier
\newpage
\fancyhead[CO,CE]{Graphs}

```{r child = '2-graphs.Rmd'}
```

\FloatBarrier
\newpage
\fancyhead[CO,CE]{Networks}

```{r child = '3-networks.Rmd'}
```

\FloatBarrier
\newpage
\fancyhead[CO,CE]{Changepoint Detection}

```{r child = '4-cpd.Rmd'}
```

\FloatBarrier
\newpage
\fancyhead[CO,CE]{Analysis}
```{r child = '5-analysis.Rmd'}
```

\FloatBarrier
\newpage
\fancyhead[CO,CE]{Conclusions}
```{r child = '6-conclusions.Rmd'}
```

\newpage
\fancyhead[CO,CE]{List of Figures}
\addcontentsline{toc}{section}{List of Figures}
\listoffigures

\newpage
\fancyhead[CO,CE]{List of Tables}
\addcontentsline{toc}{section}{List of Tables}
\listoftables

\newpage

```{r appendix_options, include=FALSE}
knitr::opts_chunk$set(fig.path = 'figures/',
                      echo = T, warning = FALSE, message = FALSE)
```

```{r child = 'appendix.Rmd'}
```

\FloatBarrier
\newpage
\fancyhead[CO,CE]{References}

# References
